On exponential sums arising from the classical doubling method

### 2025-03-04 paper.tex
Explaining notation

As usual, we rewrite this sum as follows
$$q^{-2m} \sum_{z \in \finiteFieldExtension{2m}} \sum_{y \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{x \in \NormOneGroup{2m}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(y\right)\right) \fieldCharacter_{2m}\left(z\left(y-x+1\right)\right).$$
If $z=0$ then the sum over $y$ will vanish. Hence we can reduce the sum to $z \in \multiplicativegroup{\finiteFieldExtension{2m}}$. Replacing $y$ with $z^{-1} y$ we get the sum
$$-q^{-m} \tau\left(\chi \circ \aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}, \fieldCharacter_{2m}\right) \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right)  \sum_{x \in \NormOneGroup{2m}} \theta \left(-x\right) \fieldCharacter_{2m}\left(xz\right).$$
To proceed, we use the Hilbert 90 map $\multiplicativegroup{\finiteFieldExtension{2m}} \to \NormOneGroup{2m}$ given by $t \mapsto t^{1 - q^m}$. This map is surjective, and its kernel is $\multiplicativegroup{\finiteFieldExtension{m}}$. Define $\theta' \colon \multiplicativegroup{\finiteFieldExtension{2m}} \to \multiplicativegroup{\cComplex}$ by $\theta'\left(t\right) = \theta\left(t^{1-q^m}\right)$. Replacing the sum over $x \in \NormOneGroup{2m}$ with a sum over $t \in \multiplicativegroup{\finiteFieldExtension{2m}}$, we get \begin{align*}
	& \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right) \sum_{x \in \NormOneGroup{2m}} \theta \left(x\right) \fieldCharacter_{2m}\left(xz\right) \\
	= & \frac{1}{q^m-1}\sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \fieldCharacter_{2m}\left(z^{q^m}\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t^{1-q^m} z\right).
\end{align*}
Replacing $z$ with $t^{q^m} z$, this becomes
\begin{align*}
	\frac{1}{q^m-1}\sum_{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) t\right).
\end{align*}
Since $\chi \circ \aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension} \ne \theta'$, if $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) = 0$, we have that the inner sum over $t$ is zero. Hence, we may reduce the sum over $z$ to $z$ such that $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0$. Using the fact that $\theta'$ is trivial on $\multiplicativegroup{\finiteFieldExtension{m}}$, we have after replacing variables $t \mapsto \frac{t}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right)}$,
\begin{align*}
	\frac{1}{q^m-1}\sum_{\substack{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}\\
	\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(\frac{z}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) }\right)\right) \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right).
\end{align*}
By the appendix, this equals
\begin{align*}
	q^{\frac{m}{2}} \tau\left(\chi^{-1} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right)^2 \tau\left(\chi^{2} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right) \left(\sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right)\right).
\end{align*}

@@

\begin{theorem}\label{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters}
	Suppose that $T \cong \prod_{j=1}^r \multiplicativegroup{\quadraticFieldExtension{k_j}} \times \prod_{i=1}^s \finiteFieldExtension{2m_i}^1$ is a maximal torus of $G$. Suppose that $\theta \colon T \to \multiplicativegroup{\cComplex}$ is a character and that under the above isomorphism $\theta = \alpha_1 \times \dots \times \alpha_r \times \theta_1 \times \dots \times \theta_s$, where $\alpha_j \colon \multiplicativegroup{\quadraticFieldExtension{k_j}} \to \multiplicativegroup{\cComplex}$ and $\theta_i \colon \finiteFieldExtension{2m_i}^1 \to \multiplicativegroup{\cComplex}$ are characters for every $i$ and $j$. Denote $$g_T\left(\chi, \theta, \fieldCharacter\right) = \prod_{i=1}^s \GaussSumCharacter{\theta'_i}{\chi^{-1}}{\fieldCharacter_{2m_i}} \cdot  \prod_{j=1}^r \GaussSumCharacter{\alpha_j}{\chi^{-1}}{\fieldCharacter_{k_j}} \GaussSumCharacter{\alpha_j^{-1}}{\chi^{-1}}{\fieldCharacter_{k_j}}.$$ Then
	\begin{align*}
		 \dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor} g_T\left(\chi, \theta, \fieldCharacter\right) \prod_{i=1}^s \theta_i\left(-1\right) \cdot \prod_{j=1}^r \alpha_j\left(-1\right).
	\end{align*}
\end{theorem}
\begin{proof}
	This follows from the results mentioned above and from the formula $$R_T\left(1\right) = \varepsilon_{\algebraicGroup{G}} \varepsilon_{\algebraicGroup{T}} \frac{\grpIndex{G}{T}}{\sizeof{G}_p},$$ the fact that $$ \sizeof{\lieAlgebra} = \sizeof{G}_p^2 \cdot q^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor},$$
	the facts \begin{equation*}
		\varepsilon_{\algebraicGroup{T}} = \left(-1\right)^r \text{\,\,\,\,\,\, and \,\,\,\,\,\,}  \varepsilon_{\algebraicGroup{G}} = \left(-1\right)^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor},
	\end{equation*}
	and finally from $$\GaussSumScalar{\chi^2}{\fieldCharacter} = -\tau\left(\chi^2, \fieldCharacter\right).$$
\end{proof}
\begin{remark}
	Suppose that $\pi$ is any irreducible representation of $G$ such that $\trace \pi$ appears with non-zero coefficient in $R_{T,\theta}$. Then the central character of $\pi$ satisfies
	\begin{align*}
		\centralCharacter{\pi}\left(-1\right) = \theta\left(-1\right) =& \prod_{j=1}^r \alpha_j\left(-1\right) \cdot \prod_{i=1}^s \theta_i\left(-1\right).
	\end{align*}
	Thus \begin{align*}
		\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor} \centralCharacter{\pi}\left(-1\right) g_T\left(\theta, \fieldCharacter\right).
	\end{align*}
\end{remark}

>>>>

As usual, we rewrite this sum as follows
$$q^{-2m} \sum_{z \in \finiteFieldExtension{2m}} \sum_{y \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{x \in \NormOneGroup{2m}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(y\right)\right) \fieldCharacter_{2m}\left(z\left(y-x+1\right)\right),$$
where $\fieldCharacter_{2m} = \fieldCharacter \circ \trace_{\finiteFieldExtension{2m} \slash \quadraticExtension}$

If $z=0$ then the sum over $y$ will vanish. Hence we can reduce the sum to $z \in \multiplicativegroup{\finiteFieldExtension{2m}}$. Replacing $y$ with $z^{-1} y$ we get the sum
$$-q^{-m} \tau\left(\chi \circ \aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}, \fieldCharacter_{2m}\right) \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right)  \sum_{x \in \NormOneGroup{2m}} \theta \left(-x\right) \fieldCharacter_{2m}\left(xz\right).$$
To proceed, we use the Hilbert 90 map $\multiplicativegroup{\finiteFieldExtension{2m}} \to \NormOneGroup{2m}$ given by $t \mapsto t^{1 - q^m}$. This map is surjective, and its kernel is $\multiplicativegroup{\finiteFieldExtension{m}}$. Define $\theta' \colon \multiplicativegroup{\finiteFieldExtension{2m}} \to \multiplicativegroup{\cComplex}$ by $\theta'\left(t\right) = \theta\left(t^{1-q^m}\right)$. Replacing the sum over $x \in \NormOneGroup{2m}$ with a sum over $t \in \multiplicativegroup{\finiteFieldExtension{2m}}$, we get \begin{align*}
	& \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right) \sum_{x \in \NormOneGroup{2m}} \theta \left(x\right) \fieldCharacter_{2m}\left(xz\right) \\
	= & \frac{1}{q^m-1}\sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \fieldCharacter_{2m}\left(z^{q^m}\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t^{1-q^m} z\right).
\end{align*}
Replacing $z$ with $t^{q^m} z$, this becomes
\begin{align*}
	\frac{1}{q^m-1}\sum_{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(z\right)\right) \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) t\right).
\end{align*}
Since $\chi \circ \aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension} \ne \theta'$, if $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) = 0$, we have that the inner sum over $t$ is zero. Hence, we may reduce the sum over $z$ to $z$ such that $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0$. Using the fact that $\theta'$ is trivial on $\multiplicativegroup{\finiteFieldExtension{m}}$, we have after replacing variables $t \mapsto \frac{t}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right)}$,
\begin{align*}
	\frac{1}{q^m-1}\sum_{\substack{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}\\
	\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(\frac{z}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) }\right)\right) \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right).
\end{align*}
By the appendix, this equals
\begin{align*}
	q^{\frac{m}{2}} \tau\left(\chi^{-1} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right)^2 \tau\left(\chi^{2} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right) \left(\sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \quadraticExtension}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right)\right).
\end{align*}

@@

\begin{theorem}\label{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters}
	Suppose that $T \cong \prod_{j=1}^r \multiplicativegroup{\quadraticFieldExtension{k_j}} \times \prod_{i=1}^s \finiteFieldExtension{2m_i}^1$ is a maximal torus of $G$. Suppose that $\theta \colon T \to \multiplicativegroup{\cComplex}$ is a character and that under the above isomorphism $\theta = \alpha_1 \times \dots \times \alpha_r \times \theta_1 \times \dots \times \theta_s$, where $\alpha_j \colon \multiplicativegroup{\quadraticFieldExtension{k_j}} \to \multiplicativegroup{\cComplex}$ and $\theta_i \colon \finiteFieldExtension{2m_i}^1 \to \multiplicativegroup{\cComplex}$ are characters for every $i$ and $j$. Denote $$g_T\left(\chi, \theta, \fieldCharacter\right) = \prod_{i=1}^s \GaussSumCharacter{\theta'_i}{\chi^{-1}}{\fieldCharacter_{2m_i}} \cdot  \prod_{j=1}^r \GaussSumCharacter{\alpha_j}{\chi^{-1}}{\fieldCharacter_{k_j}} \GaussSumCharacter{\alpha_j^{-1}}{\chi^{-1}}{\fieldCharacter_{k_j}},$$ where $\fieldCharacter_{2m_i} = \fieldCharacter \circ \trace_{\finiteFieldExtension{2m_i} \slash \quadraticExtension}$. Then
	\begin{align*}
		 \dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor} g_T\left(\chi, \theta, \fieldCharacter\right) \prod_{i=1}^s \theta_i\left(-1\right) \cdot \prod_{j=1}^r \alpha_j\left(-1\right).
	\end{align*}
\end{theorem}
\begin{proof}
	This follows from the results mentioned above and from the formula $$R_T\left(1\right) = \varepsilon_{\algebraicGroup{G}} \varepsilon_{\algebraicGroup{T}} \frac{\grpIndex{G}{T}}{\sizeof{G}_p},$$ the fact that $$ \sizeof{\lieAlgebra} = \sizeof{G}_p^2 \cdot q^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor},$$
	the facts \begin{equation*}
		\varepsilon_{\algebraicGroup{T}} = \left(-1\right)^r \text{\,\,\,\,\,\, and \,\,\,\,\,\,}  \varepsilon_{\algebraicGroup{G}} = \left(-1\right)^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor},
	\end{equation*}
	and finally from $$\GaussSumScalar{\chi^2}{\fieldCharacter} = -\tau\left(\chi^2, \fieldCharacter\right).$$
\end{proof}
\begin{remark}
	Suppose that $\pi$ is any irreducible representation of $G$ such that $\trace \pi$ appears with non-zero coefficient in $R_{T,\theta}$. Then the central character of $\pi$ satisfies
	\begin{align*}
		\centralCharacter{\pi}\left(-1\right) = \theta\left(-1\right) =& \prod_{j=1}^r \alpha_j\left(-1\right) \cdot \prod_{i=1}^s \theta_i\left(-1\right).
	\end{align*}
	Thus \begin{align*}
		\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim_{\quadraticExtension} \hermitianSpace}{2}\right\rfloor} \centralCharacter{\pi}\left(-1\right) g_T\left(\theta, \fieldCharacter\right).
	\end{align*}
\end{remark}
